---
title: "SARS_CoV-2 viral infection pseudotime analysis for human lung mascrophages"
output:
  html_document: default
  html_notebook: default
---

This code reproduces our infection pseudotime analysis detailed in our manuscript, "Activated interstitial macrophages are a predominant target of viral takeover and focus of inflammation in COVID-19 initiation in human lung" (https://www.biorxiv.org/content/10.1101/2022.05.10.491266v2.full).

```{r Install relevant packages}
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")

# Specify packages and their versions
packages_versions <- list(
  Seurat = "4.0.2",
  here = "1.0.1",
  sctransform = "0.3.2",
  tidyverse = "1.3.1", # This will install the tidyverse package set, which includes ggplot2, dplyr, tibble, and more.
  ggsci = "2.9",
  gplots = "3.1.1",
  heatmap3 = "1.1.9",
  PseudotimeDE = "0.9.0",
  SingleCellExperiment = "1.12.0",
  tibble = "3.1.2",
  scales = "1.1.1",
  irlba = "2.3.3"
)

# Install packages with specified versions
for (pkg in names(packages_versions)) {
  version <- packages_versions[[pkg]]
  message(sprintf("Installing %s version %s", pkg, version))
  devtools::install_version(pkg, version = version)
}
```


```{r setup, include=FALSE}
library(Seurat)
library(Matrix)
library(here)
library(sctransform)
library(loomR)
library(tidyverse)
library(SeuratDisk)
library(plyr)
library(dplyr)
library(cowplot)
library(ggsci)
library(gplots) 
library(heatmap3)

source(here::here("boilerplate_3.0.R"))
sessionInfo()
```

```{r}
suppressPackageStartupMessages(library(PseudotimeDE))
suppressPackageStartupMessages(library(SingleCellExperiment))
suppressPackageStartupMessages(library(tibble))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(irlba))
```

```{r Functions}
seurat_process <- function(seurat_object){
  
# Standard normalization 
seurat_object <- NormalizeData(seurat_object, normalization.method = "LogNormalize", scale.factor = 10000)
seurat_object <- FindVariableFeatures(seurat_object, selection.method = "vst", nfeatures = 3000)
all.genes <- rownames(seurat_object)
seurat_object <- ScaleData(seurat_object, features = all.genes)
  
  
seurat_object <- SCTransform(seurat_object, verbose = FALSE)
seurat_object <- RunPCA(seurat_object, verbose = FALSE)
return (seurat_object)
}
```

# 1. Setup seurat object and initialize
### Load seurat object of infected macrophages; seurat normalization, and PCA
```{r}
tiss.mac.1_infected_sub <- readRDS(here::here("data", "tiss.mac.1_infected.rds"))
tiss.mac.1_infected_sub <- seurat_process(tiss.mac.1_infected_sub) 
```

### Visuzlie top principal components
```{r fig.height=10, fig.width=10}
ElbowPlot(tiss.mac.1_infected_sub)
DimHeatmap(tiss.mac.1_infected_sub, dims = 1:9, cells = 500, balanced = TRUE)
DimHeatmap(tiss.mac.1_infected_sub, dims = 10:18, cells = 500, balanced = TRUE)
DimHeatmap(tiss.mac.1_infected_sub, dims = 19:27, cells = 500, balanced = TRUE)
DimPlot(tiss.mac.1_infected_sub, label = TRUE, dims = c(1,2), reduction = "pca") + NoLegend()
```

### Visualize cell clusters by UMAP 
```{r, fig.height=5, fig.width=6}
DimPlot(tiss.mac.1_infected_sub, label = TRUE, pt.size = 1.2) + NoLegend()
```

### Visualize metadata
```{r fig.height=6, fig.width=8}
UMAPPlot(tiss.mac.1_infected_sub, group.by = "tissue", label = TRUE, pt.size = 1) + ggtitle("Tissue")
UMAPPlot(tiss.mac.1_infected_sub, group.by = "cultured", label = TRUE, pt.size = 1) + ggtitle("Cultured")
UMAPPlot(tiss.mac.1_infected_sub, group.by = "condition", label = TRUE, pt.size = 1) + ggtitle("Culture Time")
UMAPPlot(tiss.mac.1_infected_sub, group.by = "infected", label = TRUE, pt.size = 1) + ggtitle("Infected")
UMAPPlot(tiss.mac.1_infected_sub, group.by = "free_annotation_MP", label = TRUE, pt.size = 1) + ggtitle("Free Annotation")
FeaturePlot(tiss.mac.1_infected_sub, features = "pct_mt_counts", pt.size = 1, cols = my.cols) + ggtitle("pct_mt_counts")
```

### Plot of viral infection pseudotime
```{r, fig.height=6, fig.width=8}
vpt_mac_mat <- tiss.mac.1_infected_sub@meta.data
ggplot(vpt_mac_mat, aes(x=log10_viral_total_UMIs, y=viral_UMIs_to_total_UMIs, color = free_annotation_MP)) +
  geom_point(size = 1) +
  geom_smooth(method = "loess", aes(colour = free_annotation_MP)) +
  theme_classic() + ylim(0,0.6) + scale_colour_manual(values = c("#0072B5FF", "#BC3C29FF"))
```

# 2. Compute viral infection pseudotime from PCA
### Visualize principal components
```{r}
tiss.mac.1_infected_sub_PCA <- tiss.mac.1_infected_sub@reductions$pca
tiss.mac.1_infected_sub_PCA_feature_loadings <- tiss.mac.1_infected_sub_PCA@feature.loadings
tiss.mac.1_infected_sub_PCA_cell_embeddings <- tiss.mac.1_infected_sub_PCA@cell.embeddings
```

### Calculate PC scores
```{r, fig.height=5, fig.width=6}
tiss.mac.1_infected_sub$PC2_3_score <- tiss.mac.1_infected_sub_PCA_cell_embeddings[,3] + tiss.mac.1_infected_sub_PCA_cell_embeddings[,2] + tiss.mac.1_infected_sub_PCA_cell_embeddings[,4] 
tiss.mac.1_infected_sub$PC2_3_score_norm = normalize(tiss.mac.1_infected_sub$PC2_3_score)

FeaturePlot(tiss.mac.1_infected_sub, features = c("PC2_3_score"), pt.size = 0.7, cols = my.cols) + ggtitle("")

tiss.mac.1_infected_sub$PC1_3score <-  tiss.mac.1_infected_sub_PCA_cell_embeddings[,1] - tiss.mac.1_infected_sub_PCA_cell_embeddings[,3]
tiss.mac.1_infected_sub$PC1score <-  tiss.mac.1_infected_sub_PCA_cell_embeddings[,1] - tiss.mac.1_infected_sub_PCA_cell_embeddings[,3] -tiss.mac.1_infected_sub_PCA_cell_embeddings[,4]

FeaturePlot(tiss.mac.1_infected_sub, features = c("PC1score"), pt.size = 0.7, cols = my.cols)

tiss.mac.1_infected_sub$PC1_3_score_norm = normalize(tiss.mac.1_infected_sub$PC1_3score)
FeaturePlot(tiss.mac.1_infected_sub, features = c("PC1_3_score_norm"), pt.size = 0.7, cols = my.cols)
```

### Visualize cell annotations, viral RNA, and PC scores
```{r, fig.height=4.5, fig.width=5}
DimPlot(tiss.mac.1_infected_sub, group.by = c("free_annotation_MP"), pt.size = 1.5, cols = rev(pal_nejm()(2)), ) + NoLegend() + ggtitle("Free Annotation") + theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) + NoLegend()

FeaturePlot(tiss.mac.1_infected_sub, features = c("log10_viral_total_UMIs"), pt.size = 1.5, cols = my.cols) + NoLegend() + ggtitle("log10_viral_UMIs") + theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) + NoLegend()

FeaturePlot(tiss.mac.1_infected_sub, features = c("PC2_3_score"), pt.size = 1.75, cols = my.cols) + NoLegend() + ggtitle("PC2_3_score") + theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) + NoLegend()

FeaturePlot(tiss.mac.1_infected_sub, features = c("PC1score"), pt.size = 1.05, cols = my.cols) + NoLegend() + ggtitle("PC1score") + theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) + NoLegend()
```

### Get PC Loadings
```{r}
# IM
PC2_3_loadings <- tiss.mac.1_infected_sub_PCA_feature_loadings[,2] + tiss.mac.1_infected_sub_PCA_feature_loadings[,3] + tiss.mac.1_infected_sub_PCA_feature_loadings[,4]
PC2_3_loadings <- as.data.frame(PC2_3_loadings)
PC2_3_loadings_genes <- rownames(PC2_3_loadings)

# AM
PC1_3_loadings <- tiss.mac.1_infected_sub_PCA_feature_loadings[,1] - tiss.mac.1_infected_sub_PCA_feature_loadings[,3] 
PC1_3_loadings <- as.data.frame(PC1_3_loadings)
PC1_3_loadings_genes <- rownames(PC1_3_loadings)
```

# 3. Perform pseudotime differential gene expression analysis
```{r}
ori_pseudotime_tbl <- tibble(cell = colnames(tiss.mac.1_infected_sub), pseudotime = tiss.mac.1_infected_sub$PC2_3_score_norm)
mac.sce <- as.SingleCellExperiment(tiss.mac.1_infected_sub)

res_fix_PC2_3 <- PseudotimeDE::runPseudotimeDE(gene.vec = PC2_3_loadings_genes[1:250],
                                     ori.tbl = ori_pseudotime_tbl,
                                     sub.tbl = NULL, # Set as NULL to only get fix.pv
                                     mat = mac.sce,
                                     model = "nb")

ori_pseudotime_AM_tbl <- tibble(cell = colnames(tiss.mac.1_infected_sub), pseudotime = tiss.mac.1_infected_sub$PC1_3_score_norm)
res_fix_PC1_3 <- PseudotimeDE::runPseudotimeDE(gene.vec = PC1_3_loadings_genes[1:250],
                                     ori.tbl = ori_pseudotime_AM_tbl,
                                     sub.tbl = NULL, # Set as NULL to only get fix.pv
                                     mat = mac.sce,
                                     model = "nb")
```